MassWateR: Improving quality control, analysis, and sharing of water quality data

### 2023-08-05 paper.Rmd
proofreading and minor edits

---
title: "MassWateR: Improving quality control, analysis, and sharing of water quality data"
date: "`r Sys.Date()`"
abstract: >
  The long-term protection and restoration of aquatic resources depends on robust monitoring data; data that require systematic quality control and analysis tools. The \CRANpkg{MassWateR} package facilitates quality control, analysis, and data sharing for discrete surface water quality data collected by monitoring programs of various size and technical capacity. The tools were developed to address regional needs for programs in Massachusetts, USA, but the principles and outputs can be applicable to monitoring data collected anywhere.  Users can create quality control reports, perform outlier analyses, and assess trends by season, date, and site for more than 40 parameters. Users can also prepare data for submission to the USEPA Water Quality Exchange, thus sharing data to the largest water quality database in the United States. The automated and reproducible workflow offered by \CRANpkg{MassWateR} is expected to increase the quantity and quality of publicly available data for the evaluation of aquatic resources.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Marcus W. Beck
    affiliation: Tampa Bay Estuary Program
    address:
    - 263 13th Ave S
    - St. Petersburg, Florida, USA 33701
    url: https://tbep.org
    orcid: 0000-0002-4996-0059
    email:  mbeck@tbep.org
  - name: Benjamen Wetherill
    url: https://www.acasak.com/
    email: bwetherill@acasak.com
    affiliation: ACASAK Consulting
    address:
    - Boston, Massachusetts, USA
  - name: Jillian Carr
    url: https://www.mass.gov/orgs/massachusetts-bays-national-estuary-partnership
    email: Jillian.Carr@umb.edu
    affiliation: Massachusetts Bays National Estuary Partnership
    address:
    - University of Massachusetts Boston, 100 Morrissey Blvd
    - Boston, Massachusetts, USA 02125
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
preamble: |
  \usepackage{multirow}
  \usepackage{multicol}
  \usepackage{colortbl}
  \usepackage{hhline}
  \newlength\Oldarrayrulewidth
  \newlength\Oldtabcolsep
  \usepackage{longtable}
  \usepackage{array}
  \usepackage{hyperref}
  \usepackage{float}
  \usepackage{wrapfig}\usepackage{longtable}
---

@@

Water quality measurements provide the foundation of environmental monitoring programs designed to protect or restore aquatic resources. In the United States, these programs are broadly guided by the federal Clean Water Act with the singular goal of restoring and maintaining the chemical, physical, and biological integrity of the nation's surface waters.  Similarly, the Water Framework Directive provides a similar legislative foundation for the protection of aquatic resources in member states of the European Union.  Numeric standards that define critical thresholds for protecting recreational, aquatic life, industrial, navigational, and consumptive uses of the resource are often established by government agencies, such that exceedances idenfitied from water quality measurements require additional regulatory action to ensure compliance. These standards and other regulatory assessments as applied at the state-level use information from long-term monitoring datasets [@Schiff16; @Tango13], or data collected *ad hoc* from multiple assessment endpoints [@Stein09; @Behmel16; @Kumpel20], where the former is atypical for most surface water bodies.  Many state or regional institutions that assess water quality rely on decentralized data sources, often combining datasets from local watershed groups or participatory science programs rather than a single database that contains adequate coverage of all areas of interest [@Buytaert14; @Kelly22]. Use of these monitoring data in a regulatory context is not possible unless standard operating procedures are adopted and the data fulfill quality control (QC) requirements.

@@

Users can engage with \CRANpkg{MassWateR} to achieve different goals.  This design was intentional based on likely differences in needs among the user community.  Although increasing data submission and facilitating QC reporting was the primary goal, we also assumed that users may not want to do both.  That is, state institutions require QC reporting for regulatory assessments, whereas data submission to WQX may be a separate process.  Users may also simply have a need to explore trends or to summarize their data, while also wanting to extend these analyses beyond \CRANpkg{MassWateR} using additional R packages. Figure \@ref(fig:workflow) demonstrates how a user may apply the functions in \CRANpkg{MassWateR} once the required data are imported.  The functions allow a user to engage with their data several ways, including 1) screening data for quality control, 2) summarizing quality control results into a single report or separate tables, 3) creating graphics for analysis and reports to stakeholders, and 4) formatting data for upload to WQX. A user can apply one to any of the four steps depending on their need.  The first step, QC screening, is often iterative as a user can modify parts of the raw data based on input checks or outliers.  The second step can be used to create a QC report for submission to a regulatory agency.  The third step can create a formatted table for WQX submission.  The fourth step is data analysis and visualization, using \CRANpkg{MassWateR} functions and downstream analysis with additional R packages and functions.  All steps require formatted input files.

@@

No matter the user need, all data inputs to \CRANpkg{MassWateR} must follow a strict format.  Developing a workflow to accommodate data inputs from the dozens of potential users from different organizations that use different data formats would have been impractical.  As such, the primary limitation to using the package is to adhere to the formatting requirements for all input files.  Several [resources](https://massbays-tech.github.io/MassWateR/RESOURCES.html) are provided on the package web page to assist users in formatting their data.  These resources included several training activities that were conducted during package development and templates demonstrating the appropriate format and rationale.  The trainings, \CRANpkg{pkdown} [website](https://massbays-tech.github.io/MassWateR/) [@Wickham22], and [Community of Practice forum](https://massbays.discourse.group/login) were implemented so that learning R was not a significant limitation for using the package.

The required data files for using \CRANpkg{MassWateR} are shown in Table \@ref(tab:filreq), including the files that apply to the workflow steps in Figure \@ref(fig:workflow).  The files are each imported into R using specific `read` functions with relevant checks, explained in the next section.  These checks verify multiple requirements outlined in the template files, with informative errors or warnings returned to the console to prompt the user on the required action to remedy a formatting issue.  The largest input file required for all parts of the workflow in Figure \@ref(fig:workflow) is the results file.  This file includes all water quality monitoring data to be used with the package and the recommendation is that the results data include a year of sampling as reporting typically follows an annual cycle. The results file design also considered WQX requirements to facilitate data submission. As such, the formatting requirements for the results file are the most burdensome for potential users and additional functions are available to assist in this effort.

@@

The primary task of the `read` functions is to ensure all imported files follow the required format for the package.  Excel files are the expected format for all inputs and the `read` functions use the `read_excel` function from the \CRANpkg{readxl} package [@Wickham23]. The `read` functions do very little other than import the file - once the file is imported it is immediately passed to one of the relevant `check` functions inside the `read` function.  There are several checks for each type of input file, with the number of checks increasing based on the complexity of the input file.  Each check is printed to the R console on completion, whereas an error is returned at the first instance of a failed check, at which point the function is exited.  The error will typically indicate which parts of the input file need to be changed to rectify the issue, often indicating a specific cell in the Excel file that requires attention. As such, the workflow is intended to be iterative, where a user imports a file, receives an error, manually changes the input file in Excel, then imports the data again until all checks pass (Figure \@ref(fig:importflow)).  Again, this design was intentional as many monitoring agencies and groups organize data differently and a standard input format for the package was the best option to accommodate all potential users.  This may also encourage future standardization among monitoring groups for how data are maintained to ease formatting challenges to using \CRANpkg{MassWateR}. A user only needs to format their data once to use the package.

```{r importflow, fig.cap = 'Pseudocode demonstrating the iterative process of importing a required data file for \\CRANpkg{MassWateR}.  All read functions import an Excel file and the imported file is then passed to a check function.  The function exits if an error is encountered, allowing the user to manually fix the identified error and then import again.  After all checks are passed, a formatting function is applied to correct minor issues (e.g., standardize date format as YYYY-MM-DD) and the final data object is returned (e.g., `resdat`).', fig.alt='Diagram showing pseudocode for importing and fixing a data file with MassWateR', out.width='100%'}
knitr::include_graphics('figs/importflow.png')
```

A correctly formatted input file would be imported as follows, with the messages in the console indicating the checks that were performed and that all checks were successful.  Below demonstrates what would be shown for the results file using an example dataset included with the package.  A total of fifteen checks are applied to the results file.

@@

A final note about the data inputs is that many types of water quality measurements can be included for analysis, although the package is currently limited to working with discrete samples as compared to data from continuous monitoring equipment.  The `paramsMWR` dataset included with \CRANpkg{MassWateR} provides a complete table of the parameters that can be used with the package.  On data import, this dataset is referenced to ensure that only relevant parameters are included and that appropriate units of measurement are provided.  The dataset includes `r nrow(paramsMWR)` different parameters, each with multiple valid units of measurement. Additionally, only one unit of measurement is allowed per parameter, which was an intentional design so that tedious functions for converting between dozens of units of measurement did not need to be created during package development.  It is not an unreasonable expectation for users to provide only one unit of measurement per parameter.  Laboratories typically have a standard reporting format based on the same methodology or instrument used to measure concentrations of water quality parameters.  Both the number of parameters and acceptable units can easily be extended and those herein were chosen simply to limit the development of \CRANpkg{MassWateR} to a manageable scope.

@@

Checking the results file for outliers is an additional and essential step of QC screening. Outliers can have many causes, including data entry errors, laboratory processing errors, or anomalous environmental events.  Identifying and potentially removing or correcting outliers is a critical part of quality control.  The functions in \CRANpkg{MassWateR} simply identify potential outliers to provide an opportunity for users to address these values.  The decision on how to address outliers remains with the user and no automated tools are provided in the package to correct or remove outliers.

The `anlzMWRoutlier()` function uses the results file and data quality objectives for accuracy to plot potential outliers for a parameter by month.  The accuracy file is used to automatically identify the y-axis scaling (as arithmetic or log) and to replace concentrations with appropriate values for those labelled as beyond detection. Outliers are defined using the standard definition of 1.5 times the interquartile range (the 25th to 75th percentile) of a parameter and can be visually identified as points above or below the whiskers in the boxplots.  The station name for a point is also shown.  The `param` argument specifies the water quality parameter to evaluate and the `group` argument specifies how the boxplots are grouped (by year, month, week, or station).  In the following example, three stations are identified as potential outliers for the respective month.

@@

A user can also evaluate outliers for every water quality parameter in the results file.  The `anlzMWRoutlierall()` function creates a Word document with images of boxplots for every parameter.  Images for each parameter can also be created as standalone files.  Here, the `param` argument does not need to be specified because all parameters are identified in the results file and processed accordingly.  The following shows how to use the function to create a single Word file or individual image files in the working directory.

@@

* *Field Blanks*: Measurements in the field to assess lack of presence of a parameter, i.e., no measurement when the parameter is not present 
* *Lab Blanks*: Laboratory samples to assess lack of presence of a parameter, i.e., no measurement when the parameter is not present
* *Field Duplicates*: Duplicate measurements in the field to assess similar values, i.e., precision between measurements is comparable
* *Lab Duplicates*: Duplicate laboratory samples to assess similar values, i.e., precision between samples is comparable
* *Calibration Checks*: Measurements in the field to verify an instrument returns a known quantity of a parameter, i.e., to verify accuracy of field sampling equipment
* *Lab Spikes*: Laboratory samples to verify if the processing equipment returns a known quantity of a parameter, i.e., to verify accuracy of laboratory equipment

Additionally, \CRANpkg{MassWateR} allows a user to include "qualified samples" in the results file.  These samples are similar as those from the above criteria and are identified by the user as having special circumstances.  For example, field duplicates are typically in pairs and the second measurement can be entered as a qualified sample in the same row of the first sample under the `"QC Reference Value"` column.  This eliminates row repetition and facilitates preparation of the results file for the user.

@@

```{r qcex, fig.cap = "The first two pages of the quality control report that evaluates the results data relative to data quality objectives.  The first page shows the data quality objectives for accuracy, frequency, and completeness.  The second page shows QC results for frequency and completeness.  Parameters shown in red or marked as 'MISS' failed the data quality objectives.  Users can edit the Word file as needed, e.g., entering the organization name or adding notes.", fig.alt = "A Word document showing the first two pages of the quality control report.", out.width = '50%', fig.show='hold'}
knitr::include_graphics('figs/qcreview1.png')
knitr::include_graphics('figs/qcreview2.png')
```

The QC report is built using several functions that can be used individually as needed.  In particular, the `tabMWRacc()`, `tabMWRfre()`, and `tabMWRcom()` create \CRANpkg{flextable} [@Gohel23] objects that can be viewed in RStudio and are compatible with Microsoft Word output.  These functions are useful for understanding how the QC checks are created for the separate components of the QC report. For example, the `tabMWRacc()` function evaluates accuracy checks for field duplicates, lab duplicates, field blanks, lab blanks, and lab spikes/instrument checks for QC records in the results file based on DQOs in the accuracy file.  The function can return a summary of all checks as follows (only ammonia and total phosphorus are shown for brevity):

@@

The cells in red showed that four of the QC checks for three of the parameters did not meet the accuracy DQOs.  All other parameters and checks in green had enough QC checks to satisfy the DQOs (note that the table colors are color-blind friendly).  Empty cells include checks where no QC records were available in the results file, e.g., no lab duplicates for dissolved oxygen.  The empty cells are typically for QC checks that do not readily apply to a parameter.  For example, dissolved oxygen is measured in the field with monitoring equipment, such that lab QC checks are not relevant.

@@

Additional functions for creating QC tables include `tabMWRfre()` for assessing if QC samples were conducted at the targeted frequency and `tabMWRcom()` for assessing completeness checks that compare the number of regular samples (field measurements or lab samples) to the number of qualified samples, i.e., a measure of the proportion of the data that is usable.  As with all the QC functions, these can be used individually within an R session to quickly view QC results or the `qcMWRreview()` function can be used to create a complete report that combines outputs from the individual functions.  The summarized report can then be submitted to an appropriate regulatory agency for review to ensure that any submitted datasets fulfill appropriate data quality objectives.  The report provides space for the user to add narrative context and notes prior to submission.

@@

The analysis functions in \CRANpkg{MassWateR} provide a streamlined approach to quickly evaluate data in the results file.  Although several base R functions and supporting packages can be used to develop custom assessments, it was recognized that many users will not be comfortable developing their own visualization routines.  As such, the analysis functions were designed for a rapid overview of the results that required minimal decisions by the user to create the output.  Several default settings described below allow further customization of the output as desired by the user. The four primary analysis functions can be used to analyze seasonal trends, trends by date, data by site, and spatially using maps.

@@

The arguments specify different options available for the plot, including which parameter to plot in the results file (`param`), the type of threshold lines to show (`thresh`), the grouping (`group`), and the type of plot (`type` as jittered points over boxplots).  The `thresh` argument, in particular, is used to plot relevant thresholds of interest based on the parameter and if the samples are from freshwater or marine sites.  The `thresholdMWR` data object included with the package includes thresholds for ten common parameters that are either regulatory standards or recommended safe limits for different designated uses in freshwater or marine environments.  Each threshold includes a source that is displayed in the legend above the plot.  In the above example, the threshold line applies to class A or B freshwater environments that support warmwater fisheries, as defined by MADEP. Threshold lines can be suppressed (`thresh = "none"`) or user-specific thresholds can de added using additional \CRANpkg{ggplot2} functionality (i.e., with `geom_hline()` added to the plot output).

@@

The `anlzMWRdate()` function plots results as time series on the date the samples or measurements were collected.  This information can be used to evaluate how samples have changed at individual sites over time or as aggregate samples across sites.  In the former case, individual points and lines are used for each site with appropriate labels. For the latter case, sites are aggregated and summary statistics are shown as the mean with 95% confidence intervals for each date.  All sites can be aggregated by sample dates or by "location groups" in the site metadata file.  For example, all sites along the same river or tributary can be aggregated.  The following demonstrates an aggregation of sites by the mainstem of the Assabet river and its tributaries using the `group = "locgroup"` argument.  A user can specify any desired grouping in the `Location Group` column of the site metadata file.

@@

Additional templates and instructions for data submission are available on the [package website](https://massbays-tech.github.io/MassWateR/index.html).  It is assumed that users are familiar with the WQX data submission portal if they are already using the package, though [training materials](https://www.epa.gov/waterdata/learn-more-about-water-quality-data) from USEPA are available.  \CRANpkg{MassWateR} eliminates the need to format the data by hand and it is expected to increase the amount of data made available on the WQP (via WQX upload) as the user base for the package increases. Further, the authors worked with USEPA WQX staff to improve WQX handling of Quality Control-coded data, by improving the list of potential field names and highlighting the need to better connect paired data (e.g., to pair lab duplicates, we created activity types "Quality Control Lab Duplicate 1" and "Quality Control Lab Duplicate 2"). Custom WQX import configurations were also developed to streamline data submission, which were made available publicly to all WQX users.

@@

To ensure that \CRANpkg{MassWateR} is a known resource for potential users and to encourage its use for QC reporting, analysis, and data submission, a community of practice was established during package development and following its initial release on CRAN (January, 2023).  This included several beta testing and training workshops to gather feedback on anticipated data analysis workflows and to educate potential users on appropriate use of the package.  Many users have not previously been exposed to R for data analysis and a substantial portion of the trainings included an [introduction](https://massbays-tech.github.io/intro-to-r/) to R, as well as \CRANpkg{MassWateR}.  Emphasis was placed on simple use of the core functions, as opposed to developing custom workflows that combined core R functions with \CRANpkg{MassWateR}.  However, users were encouraged to learn how to extend the use of the package by leveraging additional R packages, such as \CRANpkg{ggplot2} to create additional visualizations for modifying the existing analysis plots in \CRANpkg{MassWateR}.  For example, a [vignette](https://massbays-tech.github.io/MassWateR/articles/modifying.html) was included on the package website to demonstrate how these plots can be modified for custom output using \CRANpkg{ggplot2}.  Further, a [community forum](https://massbays.discourse.group/c/masswater-r-tools/5) was created as a resource for users to post questions about the package and for others to view the discussion if similar issues were encountered.  This approach followed the model used by other popular web forums (e.g., StackOverflow) for troubleshooting software issues by minimizing duplication of issues through sharing solutions in a public forum.

>>>>

---
title: "MassWateR: Improving quality control, analysis, and sharing of water quality data"
date: "`r Sys.Date()`"
abstract: >
  The long-term protection and restoration of aquatic resources depends on robust monitoring data; data that require systematic quality control and analysis tools. The \CRANpkg{MassWateR} package facilitates quality control, analysis, and data sharing for discrete surface water quality data collected by monitoring programs of various size and technical capacity. The tools were developed to address regional needs for programs in Massachusetts, USA, but the principles and outputs can be applicable to monitoring data collected anywhere.  Users can create quality control reports, perform outlier analyses, and assess trends by season, date, and site for more than 40 parameters. Users can also prepare data for submission to the USEPA Water Quality Exchange, thus sharing data to the largest water quality database in the United States. The automated and reproducible workflow offered by \CRANpkg{MassWateR} is expected to increase the quantity and quality of publicly available data to support the management of aquatic resources.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Marcus W. Beck
    affiliation: Tampa Bay Estuary Program
    address:
    - 263 13th Ave S
    - St. Petersburg, Florida, USA 33701
    url: https://tbep.org
    orcid: 0000-0002-4996-0059
    email:  mbeck@tbep.org
  - name: Benjamen Wetherill
    url: https://www.acasak.com/
    email: bwetherill@acasak.com
    affiliation: ACASAK Consulting
    address:
    - Boston, Massachusetts, USA
  - name: Jillian Carr
    url: https://www.mass.gov/orgs/massachusetts-bays-national-estuary-partnership
    email: Jillian.Carr@umb.edu
    affiliation: Massachusetts Bays National Estuary Partnership
    address:
    - University of Massachusetts Boston, 100 Morrissey Blvd
    - Boston, Massachusetts, USA 02125
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
preamble: |
  \usepackage{multirow}
  \usepackage{multicol}
  \usepackage{colortbl}
  \usepackage{hhline}
  \newlength\Oldarrayrulewidth
  \newlength\Oldtabcolsep
  \usepackage{longtable}
  \usepackage{array}
  \usepackage{hyperref}
  \usepackage{float}
  \usepackage{wrapfig}\usepackage{longtable}
---

@@

Water quality measurements provide the foundation of environmental monitoring programs designed to protect or restore aquatic resources. In the United States, these programs are broadly guided by the federal Clean Water Act with the singular goal of restoring and maintaining the chemical, physical, and biological integrity of the nation's surface waters.  Similarly, the Water Framework Directive provides a legislative foundation for the protection of aquatic resources in member states of the European Union.  Numeric standards that define critical thresholds for protecting recreational, aquatic life, industrial, navigational, and consumptive uses of the resource are often established by government agencies, such that exceedances idenfitied from water quality measurements require additional regulatory action to ensure compliance. These standards and other regulatory assessments as applied at the state-level use information from long-term monitoring datasets [@Schiff16; @Tango13], or data collected *ad hoc* from multiple assessment endpoints [@Stein09; @Behmel16; @Kumpel20], where the former is atypical for most surface water bodies.  Many state or regional institutions that assess water quality rely on decentralized data sources, often combining datasets from local watershed groups or participatory science programs rather than a single database that contains adequate coverage of all areas of interest [@Buytaert14; @Kelly22]. Use of these monitoring data in a regulatory context is not possible unless standard operating procedures are adopted and the data fulfill quality control (QC) requirements.

@@

Users can engage with \CRANpkg{MassWateR} to achieve different goals.  This design was intentional based on likely differences in needs among the user community.  Although increasing data submission and facilitating QC reporting was the primary goal, we also assumed that users may not want to do both.  That is, state institutions require QC reporting for regulatory assessments, whereas data submission to WQX may be a separate process.  Users may also simply have a need to explore trends or to summarize their data, while also wanting to extend these analyses beyond \CRANpkg{MassWateR} using additional R packages. Figure \@ref(fig:workflow) demonstrates how a user may apply the functions in \CRANpkg{MassWateR} once the required data are imported.  The functions allow a user to engage with their data several ways. The first step, QC screening, is often iterative as a user can modify parts of the raw data based on input checks or outliers. The second step can be used to create a QC report for submission to a regulatory agency. The third step is data analysis and visualization, using \CRANpkg{MassWateR} functions and downstream analysis with additional R packages and functions. The fourth and final step can create a formatted table for WQX submission.

@@

No matter the user need, all data inputs to \CRANpkg{MassWateR} must follow a strict format.  Developing a workflow to accommodate data inputs from the dozens of potential users from several organizations that use different data formats would have been impractical.  As such, the primary limitation to using the package is to adhere to the formatting requirements for all input files.  Several [resources](https://massbays-tech.github.io/MassWateR/RESOURCES.html) are provided on the package web page to assist users in formatting their data.  These resources included several training activities that were conducted during package development and templates demonstrating the appropriate format and rationale.  The trainings, \CRANpkg{pkdown} [website](https://massbays-tech.github.io/MassWateR/) [@Wickham22], and [Community of Practice forum](https://massbays.discourse.group/login) were also implemented so that learning R was not a significant limitation for using the package.

The required data files for using \CRANpkg{MassWateR} are shown in Table \@ref(tab:filreq), including how each file applies to the workflow steps in Figure \@ref(fig:workflow).  The files are imported into R using specific `read` functions with relevant checks, explained in the next section.  These checks verify multiple requirements outlined in the template files, with informative errors or warnings returned to the console to prompt the user on the required action to remedy a formatting issue.  The largest input file required for all parts of the workflow in Figure \@ref(fig:workflow) is the results file.  This file includes all water quality monitoring data to be used with the package and the recommendation is that the results data include a year of sampling as reporting typically follows an annual cycle. The results file design also considered WQX requirements to facilitate data submission. As such, the formatting requirements for the results file are the most burdensome for potential users and additional functions are available to assist in this effort.

@@

The primary task of the `read` functions is to ensure all imported files follow the required format for the package.  Excel files are the expected format for all inputs and the `read` functions use the `read_excel` function from the \CRANpkg{readxl} package [@Wickham23]. The `read` functions do very little other than import the file - once the file is imported it is immediately passed to one of the relevant `check` functions inside the `read` function.  There are several checks for each type of input file, with the number of checks increasing based on the complexity of the input file.  Each check is printed to the R console on completion, whereas an error is returned at the first instance of a failed check, at which point the function exits.  The error will typically indicate which parts of the input file need to be changed to rectify the issue, often indicating a specific cell in the Excel file that requires attention. As such, the workflow is intended to be iterative, where a user imports a file, receives an error, manually changes the input file in Excel, then imports the data again until all checks pass (Figure \@ref(fig:importflow)).  Again, this design was intentional as many monitoring agencies and groups organize data differently and a standard input format for the package was the best option to accommodate all potential users.  This may also encourage future standardization among monitoring groups for how data are maintained to ease formatting challenges to using \CRANpkg{MassWateR}. A user only needs to format their data once to use the package.

```{r importflow, fig.cap = 'Pseudocode demonstrating the iterative process of importing a required data file for \\CRANpkg{MassWateR}.  All read functions import an Excel file from the user-specified path (e.g., `respth`) and the imported file is then passed to a check function.  The function exits if an error is encountered, allowing the user to manually fix the identified error and then import again.  After all checks are passed, a formatting function is applied to correct minor issues (e.g., standardize date format as YYYY-MM-DD) and the final data object is returned (e.g., `resdat`).', fig.alt='Diagram showing pseudocode for importing and fixing a data file with MassWateR', out.width='100%'}
knitr::include_graphics('figs/importflow.png')
```

A correctly formatted input file would be imported as follows, with the messages in the console indicating the checks that were performed and that all checks were successful.  Below demonstrates what would be shown for the results file using an example dataset included with the package.  A total of fifteen checks are applied to the results file (described in detail in the [help documentation](https://massbays-tech.github.io/MassWateR/reference/checkMWRresults.html) and [vignettes](https://massbays-tech.github.io/MassWateR/articles/inputs.html#surface-water-quality-results)).

@@

A final note about the data inputs is that many types of water quality measurements can be included for analysis, although the package is currently limited to working with discrete samples as compared to data from continuous monitoring equipment.  The `paramsMWR` dataset included with \CRANpkg{MassWateR} provides a lookup table of the parameters that can be used with the package.  On data import, this dataset is referenced to ensure that only relevant parameters are included and that appropriate units of measurement are provided.  The dataset includes `r nrow(paramsMWR)` different parameters, each with multiple valid units of measurement. Additionally, only one unit of measurement is allowed per parameter, which was an intentional design so that tedious functions for converting between dozens of units of measurement did not need to be created during package development.  It is not an unreasonable expectation for users to provide only one unit of measurement per parameter.  Laboratories typically have a standard reporting format based on the same methodology or instrument used to measure water quality parameters.  Both the number of parameters and acceptable units can easily be extended and those herein were chosen simply to limit the development of \CRANpkg{MassWateR} to a manageable scope.

@@

Checking the results file for outliers is an additional and essential step of QC screening. Outliers can have many causes, including data entry errors, laboratory processing errors, or anomalous environmental events.  Identifying and potentially removing or correcting outliers is a critical part of quality control.  The functions in \CRANpkg{MassWateR} simply identify potential outliers to provide an opportunity for users to address these values.  The decision on how to address outliers remains with the user and no automated tools are provided in the package to correct or remove these values.

The `anlzMWRoutlier()` function uses the results file and data quality objectives for accuracy to plot potential outliers for a parameter by month.  The accuracy file is used to automatically identify the y-axis scaling (as arithmetic or log) and to replace concentrations with appropriate values for those labelled as beyond detection. Outliers are defined using the standard definition of 1.5 times the interquartile range (the 25th to 75th percentile) of a parameter and can be visually identified as points above or below the whiskers in the boxplots.  The station name for a point is also shown for identification.  The `param` argument specifies the water quality parameter to evaluate and the `group` argument specifies how the boxplots are grouped (by year, month, week, or station).  In the following example, three stations are identified as potential outliers for the respective month.  Note that the outliers may differ slightly by the grouping choice as they are specific to each value on the x-axis.

@@

A user can also jointly evaluate outliers for every water quality parameter in the results file.  The `anlzMWRoutlierall()` function creates a Word document with images of boxplots for every parameter.  Images for each parameter can also be created as standalone files.  Here, the `param` argument does not need to be specified because all parameters are identified in the results file and processed accordingly.  The following shows how to use the function to create a single Word file or individual image files in the working directory.

@@

* *Field Blanks*: Measurements in the field to verify absence of a parameter, i.e., the parameter is not detected by the field method when it is not present
* *Lab Blanks*: Laboratory samples to verify absence a parameter, i.e., the parameter is not detected by the lab method when it is not present
* *Field Duplicates*: Duplicate measurements in the field to assess similar values, i.e., precision between measurements is comparable
* *Lab Duplicates*: Duplicate laboratory samples to assess similar values, i.e., precision between samples is comparable
* *Calibration Checks*: Measurements in the field to verify an instrument returns a known quantity of a parameter, i.e., to verify accuracy of field sampling equipment
* *Lab Spikes*: Laboratory samples to verify if the processing equipment returns a known quantity of a parameter, i.e., to verify accuracy of laboratory equipment

Additionally, \CRANpkg{MassWateR} allows a user to include "qualified samples" in the results file.  These samples are similar as those from the above criteria and are identified by the user as having special circumstances.  For example, field duplicates are typically in pairs and the second measurement can be entered as a qualified sample in the same row of the first sample under the `"QC Reference Value"` column of the results file.  This eliminates row repetition and facilitates preparation of the results file for the user.

@@

```{r qcex, fig.cap = "The first two of sixteen pages of the quality control report created by `qcMWRreview()` that evaluates the results data relative to data quality objectives.  The first page shows the data quality objectives for accuracy, frequency, and completeness.  The second page shows QC results for frequency and completeness.  Parameters shown in red or marked as 'MISS' failed the data quality objectives.  Users can edit the Word file as needed, e.g., entering the organization name or adding notes.", fig.alt = "A Word document showing the first two pages of the quality control report.", out.width = '50%', fig.show='hold'}
knitr::include_graphics('figs/qcreview1.png')
knitr::include_graphics('figs/qcreview2.png')
```

The QC report is built using several functions that can be used individually as needed.  In particular, the `tabMWRacc()`, `tabMWRfre()`, and `tabMWRcom()` create \CRANpkg{flextable} [@Gohel23] objects that can be viewed in RStudio and are compatible with Word output.  These functions are useful for understanding how the QC checks are created for the separate components of the QC report. For example, the `tabMWRacc()` function evaluates accuracy checks for field duplicates, lab duplicates, field blanks, lab blanks, and lab spikes/instrument checks for QC records in the results file based on DQOs in the accuracy file.  The function can return a summary of all checks as follows (only ammonia and total phosphorus are shown for brevity):

@@

The cells in red show that four of the QC checks for three of the parameters did not meet the accuracy DQOs.  All other parameters and checks in green had enough QC checks to satisfy the DQOs (note that the table colors are color-blind friendly).  Empty cells include checks where no QC records were available in the results file, e.g., no lab duplicates for dissolved oxygen.  The empty cells are typically for QC checks that do not readily apply to a parameter.  For example, dissolved oxygen is measured in the field with monitoring equipment, such that lab QC checks are not relevant.

@@

Additional functions for creating QC tables include `tabMWRfre()` for assessing if QC samples were conducted at the targeted frequency and `tabMWRcom()` for assessing completeness checks that compare the number of regular samples (field measurements or lab samples) to the number of qualified samples, i.e., a measure of the proportion of the data that is usable.  As with all the QC functions, these can be used individually within an R session to quickly view QC results or the `qcMWRreview()` function can be used to create a complete report that combines outputs from the individual functions.  The summarized report can then be submitted to an appropriate regulatory agency for review to ensure that any submitted datasets fulfill appropriate data quality objectives.

@@

The analysis functions in \CRANpkg{MassWateR} provide a streamlined approach to quickly evaluate data in the results file.  Although several base R functions and supporting packages can be used to develop custom assessments, it was recognized that many users will not be comfortable developing their own visualization routines.  As such, the analysis functions were designed for a rapid overview of the results that require minimal decisions by the user to create the output.  Several default settings described below allow further customization of the output as desired by the user. The four primary analysis functions can be used to analyze seasonal trends, trends by date, data by site, and spatially using maps.

@@

The arguments specify different options available for the plot, including which parameter to plot in the results file (`param`), the type of threshold line(s) to show (`thresh`), the grouping (`group`), and the type of plot (`type` as jittered points over boxplots).  The `thresh` argument, in particular, is used to plot relevant thresholds of interest based on the parameter and if the samples are from freshwater or marine sites.  The `thresholdMWR` data object included with the package includes thresholds for ten common parameters that are either regulatory standards or recommended safe limits for different designated uses in freshwater or marine environments.  Each threshold includes a source that is displayed in the legend above the plot.  In the above example, the threshold line applies to class A or B freshwater environments that support warmwater fisheries, as defined by MADEP. Threshold lines can be suppressed (`thresh = "none"`) or user-specific thresholds can de added using additional \CRANpkg{ggplot2} functionality (i.e., with `geom_hline()` added to the plot output).

@@

The `anlzMWRdate()` function plots results as time series on the date the samples or measurements were collected.  This information can be used to evaluate how samples have changed at individual sites over time or as aggregate samples across sites.  In the former case, individual points and lines are used for each site with appropriate labels. For the latter case, sites are aggregated and summary statistics are shown as the mean with optional 95% confidence intervals for each date.  All sites can be aggregated by sample dates or by "location groups" in the site metadata file.  For example, all sites along the same river or tributary can be aggregated.  The following demonstrates an aggregation of sites by the mainstem of the Assabet river and its tributaries using the `group = "locgroup"` argument.  A user can specify any desired grouping in the `Location Group` column of the site metadata file.

@@

Additional templates and instructions for data submission are available on the [package website](https://massbays-tech.github.io/MassWateR/index.html).  It is assumed that users are familiar with the WQX data submission portal if they are already using the package, though [training materials](https://www.epa.gov/waterdata/learn-more-about-water-quality-data) from USEPA are available.  \CRANpkg{MassWateR} eliminates the need to format the data by hand and it is expected to increase the amount of data made available on the WQP (via WQX upload) as the user base for the package increases. Further, the authors worked with USEPA WQX staff to improve WQX handling of Quality Control-coded data, by improving the list of potential field names and highlighting the need to better connect paired data (e.g., activity types "Quality Control Lab Duplicate 1" and "Quality Control Lab Duplicate 2" were created to pair lab duplicates). Custom WQX import configurations were also developed to streamline data submission, which were made available publicly to all WQX users.

@@

To ensure that \CRANpkg{MassWateR} is a known resource for potential users and to encourage its use for QC reporting, analysis, and data submission, a community of practice was established during package development and following its initial release on CRAN (January, 2023).  This included several beta testing and training workshops to gather feedback on anticipated data analysis workflows and to educate potential users on appropriate use of the package.  Many users have not previously been exposed to R for data analysis and a substantial portion of the trainings included an [introduction](https://massbays-tech.github.io/intro-to-r/) to R, as well as \CRANpkg{MassWateR}.  Emphasis was placed on simple use of the core functions, as opposed to developing custom workflows that combined core R functions with \CRANpkg{MassWateR}.  However, users were encouraged to learn how to extend the use of the package by leveraging additional R packages, such as \CRANpkg{ggplot2} to create additional visualizations for modifying the existing analysis plots in \CRANpkg{MassWateR}.  For example, a [vignette](https://massbays-tech.github.io/MassWateR/articles/modifying.html) was included on the package website to demonstrate how these plots can be modified for custom output using \CRANpkg{ggplot2}.  Further, a [community forum](https://massbays.discourse.group/c/masswater-r-tools/5) was created as a resource for users to post questions about the package and for others to view the discussion if similar issues were encountered.  This approach followed the model used by other popular web forums (e.g., StackOverflow) for troubleshooting software issues by minimizing duplication of topics through sharing solutions in a public forum.
